---
title: "Class 5: Data Viz with ggplot"
author: "Melanie Alonzo (A17375327)"
format: pdf
toc: true
---

## Background

There are lots of ways to make plots in R. These include so-called "base R" (like the `plot()`) and add on packages like **ggplot 2**.

Lets make the same plot with these two graphics systems. We can use the inbuilt `cars` dataset:

```{r}
head ( cars )
```

With "Base R" we can simply:

```{r}
plot(cars)
```

Now lets try ggplot. First I need to install the package using `install.packages("ggplot2")`. 

> **N.B.** We never run an `install.packages()` in a code chunk otherwise we will re-install needlesely every time we render our document.

Every time we want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least 3 things:

1. The **data** i.e. stuff to plot as a data.frame
2. The **aes** or aesthetics that map the data to the plot
3. The **geom_** or geometry i.e. the plot type such as points, lines etc.


```{r}
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x="speed(MPH)", y="distance(ft)", title="Stopping Distance of old Cars")+
  theme_bw()
```


## Gene Expression Plot

Read some data on the effects of GLP-1 inhibitor (drug) on gene expression values:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page.

```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2)+
  geom_point(col="blue", alpha=0.2)
```

Let's color by `state` up, down or no change.

```{r}
table( genes$State )
```

```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()+
  scale_color_manual(values= c("brown", 
                               "gray", 
                               "orange"))+
  labs(x= "control (no drug)", 
       y= "Drug Treatment", 
       title= "Expression changes with. GLP-1 drug")+
  theme_minimal()
```

## Going further with gapminder

Here we explore the famous `gapminder` dataset with some custom plots.

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many rows does this dataset have?

```{r}
nrow(gapminder)
```

> Q. How many different continents are in this dataset?

```{r}
table( gapminder$continent )
```

Version 1 plot GDP vs LifeExp for all rows

```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()
```
I want to see a plot for each continent - in ggplot lingo this is called "faceting"

```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()+
  facet_wrap(~continent)
```

## First look at the dplyer package

Another add-on package with a function called `filter()` that we want to use.

```{r}
library(dplyr)
```

```{r}
head(gapminder)
```


```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
input <- filter(gapminder, year==2007 | year==1977)
```

```{r}
ggplot(input)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()+
  facet_wrap(~year)
```

